---
title: "DataDescription"
author: "Jiayin Liu"
format: html
editor: visual
---

```{r, echo=FALSE,warning=FALSE}
library(readxl)
library(tidyverse)
library(dplyr)
library(janitor)
library(scales)
library(lubridate)
library(ggExtra)
library(ggplot2)
library(ggridges)
```

# Data preparation

**import the data**

```{r,warning=FALSE}
dt <- readxl::read_excel("US_Superstore_data.xls")
```

**clean the data** 

step 1: rename the colnames that can cause problem

```{r}
dt_col <- dt %>%
  clean_names()
colnames(dt)
```

step 2: check the data type and make the variable to the class I want

```{r}
str(dt_col)
dt_date <- dt_col %>%
  mutate(order_date = as.Date(order_date),
         ship_date = as.Date(ship_date),
         order_year = year(order_date),
         order_month = month(order_date,label = TRUE,abbr = TRUE,locale = "C"),# I do this locale=C because my pc speak Chinese 
         order_year_month = floor_date(order_date,unit = "month"),
         shipping_days = as.numeric(ship_date - order_date),
         profit_status = case_when(profit > 0 ~ "Profit",
                                   profit == 0 ~ "Break Even",
                                   profit < 0 ~ "Loss",
                                   TRUE ~ NA))
```

**check the data structure**

```{r}
str(dt_date)

```

**check the missing data**

```{r}
dt_date %>%summarise(across(everything(),~ sum(is.na(.))))->missing_summary # 
missing_summary
```

Based on the summary output. seems no missing items under each variable. Perfect!

# Contingency tables

## one way contingency table

```{r}
#1. product category
category_table <- dt_date %>%
  count(category) %>%
  mutate(percent = n / sum(n))
category_table

#2. customer segment
segment_table <- dt_date %>%
  count(segment) %>%
  mutate(percent = n / sum(n))
segment_table

#3. region
region_table <- dt_date %>%
  count(region) %>%
  mutate(percent = n / sum(n))
region_table

#4. profit status
profit_status_table <- dt_date %>%
  count(profit_status) %>%
  mutate(percent = n / sum(n))
profit_status_table

```

## two way contingency table

```{r}
#1. category and segment
category_segment_table <- dt_date %>%
  drop_na(category, segment) %>%
  group_by(category, segment) %>%
  summarize(count = n(),.groups = "drop") %>%
  pivot_wider(names_from = segment,values_from = count,values_fill = 0)
category_segment_table

#2. region and profit status
region_profit_table <- dt_date %>%
  drop_na(region, profit_status) %>%
  group_by(region, profit_status) %>%
  summarize(count = n(),.groups = "drop") %>%
  pivot_wider(names_from = profit_status,values_from = count,values_fill = 0)
region_profit_table


#3. ship mode by customer segment
category_segment_table <- dt_date %>%
  drop_na(category, segment) %>%
  group_by(category, segment) %>%
  summarize(count = n(),.groups = "drop") %>%
  pivot_wider(names_from = segment,values_from = count,values_fill = 0)
category_segment_table

#4. category by region
category_region_table <- dt_date %>%
  drop_na(category, region) %>%
  group_by(category, region) %>%
  summarize(count = n(),.groups = "drop") %>%
  pivot_wider(names_from = region,values_from = count,values_fill = 0)
category_region_table

```

# Numerical summary

```{r}
#1. overall numerical summary
overall_numeric_summary <- dt_date %>%
  summarize(across(c(sales,quantity,discount,profit,shipping_days),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        median = ~ median(.x, na.rm = TRUE),
                        sd = ~ sd(.x, na.rm = TRUE),
                        IQR = ~ IQR(.x, na.rm = TRUE)),.names = "{.col}_{.fn}")) %>%
  pivot_longer(cols = everything(),
               names_to = c("variable", "statistic"),
               names_pattern = "^(.*)_(mean|median|sd|IQR)$",
               values_to = "value") %>%
  pivot_wider(names_from = statistic,values_from = value)

overall_numeric_summary

#2.numerical summary by product category
numeric_summary_by_category <- dt_date %>%
  group_by(category) %>%
  drop_na(category) %>%
  summarize(
    across(c(sales,quantity,discount,profit,shipping_days),
           list(mean = ~ mean(.x, na.rm = TRUE),
                median = ~ median(.x, na.rm = TRUE),
                sd = ~ sd(.x, na.rm = TRUE),
                IQR = ~ IQR(.x, na.rm = TRUE)),
           .names = "{.fn}_{.col}"))
numeric_summary_by_category

#3. numerical summary by customer segment
numeric_summary_by_segment <- dt_date %>%
  group_by(segment) %>%
  drop_na(segment) %>%
  summarize(
    across(c(sales,quantity,discount,profit,shipping_days),
           list(mean = ~ mean(.x, na.rm = TRUE),
                median = ~ median(.x, na.rm = TRUE),
                sd = ~ sd(.x, na.rm = TRUE),
                IQR = ~ IQR(.x, na.rm = TRUE)),
           .names = "{.fn}_{.col}"))
numeric_summary_by_segment

#4. numerical summary by region
numeric_summary_by_region <- dt_date %>%
  group_by(region) %>%
  drop_na(region) %>%
summarize(
    across(c(sales,quantity,discount,profit,shipping_days),
           list(mean = ~ mean(.x, na.rm = TRUE),
                median = ~ median(.x, na.rm = TRUE),
                sd = ~ sd(.x, na.rm = TRUE),
                IQR = ~ IQR(.x, na.rm = TRUE)),
           .names = "{.fn}_{.col}"))
numeric_summary_by_region

#5. numerical summary by customer segment
numeric_summary_by_category_segment <- dt_date %>%
  group_by(category, segment) %>%
  drop_na(category, segment) %>%
summarize(
    across(c(sales,quantity,discount,profit,shipping_days),
           list(mean = ~ mean(.x, na.rm = TRUE),
                median = ~ median(.x, na.rm = TRUE),
                sd = ~ sd(.x, na.rm = TRUE),
                IQR = ~ IQR(.x, na.rm = TRUE)),
           .names = "{.fn}_{.col}"))
numeric_summary_by_category_segment
```

# Plot

For the regular plots:

```{r}
#1. bar chart of product categories
plot1 <- ggplot(dt_date,aes(x = category)) +
  geom_bar() +
  labs(
    title = "Number of Order Items by Product Category",
    subtitle = "Frequency of records in each product category",
    x = "Product Category",
    y = "Number of Order Items"
  ) +
  theme_minimal()

plot1

#2.Side-by-side bar chart of category and customer segment
plot2 <- ggplot(dt_date,aes(x = category,fill = segment)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Order Items by Product Category and Customer Segment",
    subtitle = "Comparison of customer purchasing patterns across categories",
    x = "Product Category",
    y = "Number of Order Items",
    fill = "Customer Segment"
  ) +
  theme_minimal()

plot2

#3. Histogram of sales
plot3 <- dt_date %>%
  filter(sales > 0,!is.na(sales)) %>%
  ggplot(aes(x = sales)) +
  geom_histogram(bins = 40,color = "white" ) +
  scale_x_log10(labels = dollar ) +
  labs(
    title = "Distribution of Sales",
    subtitle = "Sales are displayed on a logarithmic scale",
    x = "Sales",
    y = "Number of Order Items"
  ) +
  theme_minimal()

plot3

#4.Box plot of profit by category and customer segment
profit_lower <- quantile(dt_date$profit,0.01,na.rm = TRUE)

profit_upper <- quantile(dt_date$profit,0.99,na.rm = TRUE)

plot4 <- ggplot(dt_date,aes(x = category,y = profit,fill = segment)) +
  geom_boxplot(outlier.alpha = 0.2,
               position = position_dodge(width = 0.8)) +
  coord_cartesian(ylim = c(profit_lower,profit_upper) ) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Profit Distribution by Product Category and Customer Segment",
    subtitle = "The vertical axis displays the 1st through 99th percentiles",
    x = "Product Category",
    y = "Profit",
    fill = "Customer Segment"
  ) +
  theme_minimal()

plot4

#5. Scatterplot of sales and profit
plot5 <- dt_date %>%
  filter(sales > 0,!is.na(profit),!is.na(category),!is.na(quantity)) %>%
  ggplot(aes(x = sales,y = profit,color = category,size = quantity)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0,linetype = "dashed") +
  scale_x_log10(labels = dollar ) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Relationship Between Sales and Profit",
    subtitle = "Color represents category and point size represents quantity",
    x = "Sales on a Logarithmic Scale",
    y = "Profit",
    color = "Product Category",
    size = "Quantity"
  ) +
  theme_minimal()

plot5
```

Extra plot not learned from the class

**plot6:** This plot shows whether higher discounts are associated with lower profits. Also shows whether that relationship differs across product categories and customer segments.

```{r,warning=FALSE}
#6. Discount and profit by product category
plot6 <- dt_date %>%
  drop_na(discount,profit,category,segment ) %>%
  ggplot(aes(x = discount,y = profit,color = segment)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess",se = FALSE,linewidth = 0.8) +
  geom_hline(yintercept = 0,linetype = "dashed") +
  facet_wrap(~ category ) +
  coord_cartesian(ylim = c(profit_lower,profit_upper ) ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title = "Relationship Between Discount and Profit by Product Category",
    subtitle = "Smooth curves show the general pattern for each customer segment",
    x = "Discount",
    y = "Profit",
    color = "Customer Segment"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

plot6
```

**plot 7:** This plot shows how sales change over time. In addition, the plot also tells me whether the three product categories follow similar or different patterns.

```{r}
#7. Monthly sales trends by product category
monthly_category_sales <- dt_date %>%
  drop_na(order_year_month,category,sales) %>%
  group_by(order_year_month,category ) %>%
  summarize(total_sales = sum(sales,na.rm = TRUE),.groups = "drop")

monthly_category_sales

plot7 <- ggplot(
  monthly_category_sales,
  aes(x = order_year_month,y = total_sales,color = category)) +
  geom_line(linewidth = 0.9 ) +
  geom_point(size = 1.2 ) +
  scale_x_date(date_breaks = "6 months",
               labels = scales::label_date(format = "%b %Y",locale = "C")) +
  scale_y_continuous(labels = scales::dollar ) +
  labs(
    title = "Monthly Sales Trends by Product Category",
    subtitle = "Total sales are aggregated by month",
    x = "Order Month",
    y = "Total Monthly Sales",
    color = "Product Category") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45,hjust = 1))

plot7

```

**plot8:** This ridgeline plot compares the full profit distributions across product sub-categories.

```{r}
#8.Profit distributions across product sub-categories
plot8 <- dt_date %>%
  drop_na(sub_category,profit) %>%
  dplyr::filter(profit >= profit_lower,profit <= profit_upper ) %>%
  ggplot(aes(x = profit,
             y = forcats::fct_reorder(sub_category,profit,median),
             fill = after_stat(x)) ) +
  geom_density_ridges_gradient(scale = 2.5,rel_min_height = 0.01 ) +
  geom_vline(xintercept = 0,linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar) +
  scale_fill_viridis_c(labels = scales::dollar ) +
  labs(
    title = "Profit Distributions Across Product Sub-Categories",
    subtitle = "Sub-categories are ordered by median profit",
    x = "Profit",
    y = "Product Sub-Category",
    fill = "Profit"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot8

```

**plot 9:** This plot shows the relationship between sales and profit. It also shows the separate distributions of both variables.

```{r}

#9.Sales and profit with marginal distributions
base_plot <- dt_date %>%
  dplyr::filter(sales > 0,!is.na(profit),!is.na(category)) %>%
  ggplot(aes(x = log10(sales),y = profit,color = category)) +
  geom_point(alpha = 0.35) +
  geom_hline(yintercept = 0,linetype = "dashed") +
  coord_cartesian(ylim = c(profit_lower,profit_upper) ) +
  labs(
    title = "Sales and Profit with Marginal Distributions",
    subtitle = "Marginal density curves show the distributions of sales and profit",
    x = "Log10 Sales",
    y = "Profit",
    color = "Product Category"
  ) +
  theme_minimal() +
  theme(legend.position = "right" )

plot9 <- ggExtra::ggMarginal(
  base_plot,
  type = "density",
  groupColour = TRUE,
  groupFill = TRUE,
  alpha = 0.3
)

plot9
```
